---
title: "Quick Plots"
author: NULL
date: NULL
output:
  html_document:
    keep_md: true
    fig_width: 4
    fig_height: 4
---

For quick data exploration, base R plotting functions can provide an expeditious and straightforward approach to understanding your data. These functions are installed by default in base R and do not require additional visualization packages to be installed. This straightforward tutorial should teach you the basics, and give you a good idea of what you want to do next.  

In addition, I'll show how to make similar graphics with the `qplot()` function in `ggplot2`, which has a syntax similar to the base graphics functions. For each `qplot()` graph, there is also an equivalent using the more powerful `ggplot()` function which I illustrate in later visualization tutorials. This will, hopefully, help you transition to using `ggplot2` when you want to make more sophisticated graphics. 

<br>

## tl;dr
Don't have the time to scroll through the full tutorial?  Skip directly to the section of interest:

- [Replication requirements](#replication)
- [Scatter plot](#scatter)
- [Line chart](#line)
- [Bar chart](#bar)
- [Histogram](#histogram)
- [Box plot](#box)
- [Stem & leaf plot](#stem)

<br>

## Replication Requirements {#replication}
To illustrate these quick plots I'll use several built in data sets that come with base R. R has 104 built in data sets that can be viewed with `data()`. The ones I'll use below include `mtcars`, `pressure`, `BOD`, and `faithful`. You can type these in your R console at anytime to see the data.  Also, in addition to base R plotting functions I illustrate how to use the `qplot()` function from the `ggplot2` package.

```{r, eval=FALSE, collapse=TRUE}
# data sets used
mtcars
pressure
BOD
faithful

# package used
library(ggplot2)
```

&#9755; *See [Working with packages](http://bradleyboehmke.github.io/tutorials/basics/packages/) for more information on installing, loading, and getting help with packages.*

```{r, message=FALSE, warning=FALSE, echo=FALSE}
library(ggplot2)
theme_update(plot.background = element_rect(fill = "#fdfdfd"))
```

<br>

## Scatter Plot {#scatter}
To make a scatter plot use `plot()` with a vector of x values and a vector of y values:

```{r, echo=TRUE, eval=FALSE}
# base R
plot(x = mtcars$wt, y = mtcars$mpg)
```

```{r, echo=FALSE, fig.align='center'}
# base R
par(bg = "#fdfdfd")
plot(x = mtcars$wt, y = mtcars$mpg)
```

You can get a similar result using `qplot()`:

```{r, echo=TRUE, fig.align='center'}
library(ggplot2)
qplot(x = mtcars$wt, y = mtcars$mpg)
```

If the two vectors are already in the same data frame, note that the following functions produce the same outpt:

```{r, echo=TRUE, eval=FALSE}

# specifying only x and y vectors
qplot(x = mtcars$wt, y = mtcars$mpg)

# specifying x and y vectors from a data frame
qplot(x = wt, y = mpg, data = mtcars)

# using full ggplot syntax
ggplot(data = mtcars, aes(x = wt, y = mpg)) +
        geom_point()
```

You can also get a scatter plot matrix to observe several plots at once.  In this case you just pass the multiple variables (columns) in the data frame to `plot()` and a scatter plot matrix will be returned. The `qplot()` function does not have this same functionality; however, you can do more advanced plotting matrices by using `ggplot()`'s facetting arguments.  This will be covered in later tutorials.

```{r, echo=TRUE, eval=FALSE}
# passing multiple variables to plot
plot(mtcars[, 4:6])
```

```{r, echo=FALSE, fig.align='center'}
# passing multiple variables to plot
par(bg = "#fdfdfd")
plot(mtcars[, 4:6])
```

<a href="#top">Go to top</a>

<br>

## Line Chart {#line}
By default the `plot()` function produces a scatter plot with dots. To make a line graph, pass it the vector of x and y values, and specify `type = "l"` for *line*:

```{r, echo=TRUE, eval=FALSE}
plot(x = pressure$temperature, y = pressure$pressure, type = "l")
```

```{r, echo=FALSE, fig.align='center'}
par(bg = "#fdfdfd")
plot(x = pressure$temperature, y = pressure$pressure, type = "l")
```

Similarly, you can pass it the argument `type = "s"` to produce a stepped line chart:
```{r, echo=TRUE, eval=FALSE}
plot(x = pressure$temperature, y = pressure$pressure, type = "s")
```

```{r, echo=FALSE, fig.align='center'}
par(bg = "#fdfdfd")
plot(x = pressure$temperature, y = pressure$pressure, type = "s")
```

To include multiple lines or to plot the points, first call `plot()` for the first line, then add additional lines and points with `lines()` and `points()` respectively: 

```{r, echo=TRUE, eval=FALSE}
# base graphic
plot(x = pressure$temperature, y = pressure$pressure, type = "l")

# add points
points(x = pressure$temperature, y = pressure$pressure)

# add second line in red color
lines(x = pressure$temperature, y = pressure$pressure/2, col = "red")

# add points to second line
points(x = pressure$temperature, y = pressure$pressure/2, col = "red")
```

```{r, echo=FALSE, fig.align='center'}
par(bg = "#fdfdfd")
# base graphic
plot(x = pressure$temperature, y = pressure$pressure, type = "l")

# add points
points(x = pressure$temperature, y = pressure$pressure)

# add second line in red color
lines(x = pressure$temperature, y = pressure$pressure/2, col = "red")

# add points to second line
points(x = pressure$temperature, y = pressure$pressure/2, col = "red")
```

We can use `qplot()` to get similar results by using the `geom` argument. `geom` means adding a geometric object (line, points, etc.) to visually represent the data and in this case we want to represent the data using a line and then also points:

```{r, echo=TRUE, fig.align='center'}
# using qplot for a line chart
qplot(temperature, pressure, data = pressure, geom = "line")

# using qplot for a stepped line chart
qplot(temperature, pressure, data = pressure, geom = "step")

# using qplot for a line chart with points
qplot(temperature, pressure, data = pressure, geom = c("line", "point"))
```

We can get the same output using the full `ggplot()` syntax:
```{r, echo=TRUE, eval=FALSE}
# line chart
ggplot(pressure, aes(x = temperature, y = pressure)) +
        geom_line()

# step chart
ggplot(pressure, aes(x = temperature, y = pressure)) +
        geom_step()

# line chart with points
ggplot(pressure, aes(x = temperature, y = pressure)) +
        geom_line() +
        geom_point()
```


<a href="#top">Go to top</a>

<br>

## Bar Chart {#bar}
To make a bar chart of values, use `barplot()` and pass it a vector of values for the height of each bar and (optionally) a vector of labels for each bar. If the vector has names for the elements, the names will automatically be used as labels:

```{r, echo=TRUE, eval=FALSE}
barplot(height = BOD$demand, names.arg = BOD$Time)
```

```{r, echo=FALSE, fig.align='center'}
par(bg = "#fdfdfd")
barplot(height = BOD$demand, names.arg = BOD$Time)
```

When you want the bar chart to represent the *count* of cases in each category then you need to generate the count of unique values. For instance, in the `mtcars` dataset we may want to look at the cylinder variable and understand the distribution. To do this we can use the `table()` function which will provide us the count of each unique value in this variable.  We can then pass this to the `barplot()` function to plot the counts of cylinders:

```{r, echo=TRUE, collapse=TRUE, eval=FALSE}
# the cylinder variable in the mtcars dataset is made up of values of 4, 6 & 8
mtcars$cyl
##  [1] 6 6 4 6 8 6 8 4 4 6 6 8 8 8 8 8 8 4 4 4 4 8 8 8 8 4 4 4 8 6 8 4

# get the count of 4, 6 & 8 cylinder cars in the dataset
table(mtcars$cyl)
## 
##  4  6  8 
## 11  7 14

# plot the count of 4, 6 & 8 cylinder cars in the dataset
barplot(table(mtcars$cyl))
```

```{r, echo=FALSE, fig.align='center', collapse=TRUE}
# plot the count of 4, 6 & 8 cylinder cars in the dataset
par(bg = "#fdfdfd")
barplot(table(mtcars$cyl))
```

To get the same result using `qplot()` we use `geom = "bar"`.

```{r, echo=TRUE, fig.align='center'}
# x defaults to a continuous variable
qplot(mtcars$cyl, geom = "bar")
```

Note how the x axis defaults to a continuous variable in the plot above. Since bar charts are designed for categorical variables we want our x variable to a factor variable so that our x axis appropriately represents the data.  

```{r, echo=TRUE, fig.align='center'}
# use factor(x) to make it discrete
qplot(factor(mtcars$cyl), geom = "bar")
```

&#9755; *See the [Factors](http://bradleyboehmke.github.io/tutorials/factors) tutorial for more information on categorical variables (aka factors) in R.*

<a href="#top">Go to top</a>

<br>

## Histogram {#histogram}
To make a histogram, use `hist()` and pass it a single vector of values. You can also use the `breaks` argument to determine the size of the bins.

```{r, echo=TRUE, eval=FALSE}
# default bins
hist(mtcars$mpg)
```

```{r, echo=FALSE, fig.align='center'}
# default bins
par(bg = "#fdfdfd")
hist(mtcars$mpg)
```

```{r, echo=TRUE, eval=FALSE}
# adjust binning
hist(mtcars$mpg, breaks = 10)
```

```{r, echo=FALSE, fig.align='center'}
# adjust binning
par(bg = "#fdfdfd")
hist(mtcars$mpg, breaks = 10)
```

To get the same result using `qplot()` we don't need to specify a `geom` argument because when you feed `qplot()` with a single variable it will default to using a histogram. You can also control the binning by using the `binwidth` argument. Although not necessary I add the `color` argument to outline the bars.

```{r, echo=TRUE, fig.align='center'}
qplot(mtcars$mpg, binwidth = 3, color = I("white"))
```

<a href="#top">Go to top</a>

<br>

## Box Plot {#box}
To make a box-whisker plot (aka box plot), use plot() and pass it x values that are categorical (aka factor) and a vector of y values. However, you need to ensure that the x values are factors otherwise you will get a scatter plot by default:

```{r, echo=TRUE, eval=FALSE}
# if x is not a factor it will produce a scatter plot
plot(x = mtcars$cyl, y = mtcars$mpg)
```

```{r, echo=FALSE, fig.align='center'}
# if x is not a factor it will produce a scatter plot
par(bg = "#fdfdfd")
plot(mtcars$cyl, mtcars$mpg)
```

When x is a [factor](http://bradleyboehmke.github.io/tutorials/factors) (as opposed to a numeric vector), it will automatically create a box plot:

```{r, echo=TRUE, eval=FALSE}
# if x is a factor it will produce a box plot
plot(factor(mtcars$cyl), mtcars$mpg)
```

```{r, echo=FALSE, fig.align='center'}
# if x is a factor it will produce a box plot
par(bg = "#fdfdfd")
plot(factor(mtcars$cyl), mtcars$mpg)
```



Alternatively, we can use the `boxplot()` function to create a box plot.  We can create a single box plot with the following:

```{r, echo=TRUE, eval=FALSE}
# boxplot of mpg
boxplot(mtcars$mpg)
```

```{r, echo=FALSE, fig.align='center'}
# boxplot of mpg
par(bg = "#fdfdfd")
boxplot(mtcars$mpg)
```

To get a box plot that displays the distribution of mpg values across the different cylinders we use the "~" to state that we want to assess *y* by *x*:  

```{r, echo=TRUE, eval=FALSE}
# boxplot of mpg by cyl
boxplot(mpg ~ cyl, data = mtcars)
```

```{r, echo=FALSE, fig.align='center'}
# boxplot of mpg by cyl
par(bg = "#fdfdfd")
boxplot(mpg ~ cyl, data = mtcars)
```

We can also assess interactions. In this case we look at the distribution of mpg by cylinders and transmission. Note on the y axis is mpg and on the x axis are the cylinder ~ transmission interaction. Note that the transmission variable is coded as 0 for automatic and 1 for manual. So the x-axis values of 4.0, 6.0, 8.0, 4.1, etc. represent 4 cylinder with automatic transmission, 6 cylinder with automatic transmission, 8 cylinder with automatic transmission, 4 cylinder with manual transmission, etc. 

```{r, echo=TRUE, eval=FALSE}
# boxplot of mpg based on interaction of two variables
boxplot(mpg ~ cyl + am, data = mtcars)
```

```{r, echo=FALSE, fig.align='center'}
# boxplot of mpg based on interaction of two variables
par(bg = "#fdfdfd")
boxplot(mpg ~ cyl + am, data = mtcars)
```

Similar results are attained with `qplot()` using `geom = "boxplot"`:

```{r, echo=TRUE, fig.align='center'}
qplot(x = factor(cyl), y = mpg, data = mtcars, geom = "boxplot")
```

<a href="#top">Go to top</a>

<br>

## Stem & Leaf Plot {#stem}
To make a stem-and-leaf plot we can simply use the `stem()` function and pass it a vector of numeric values:

```{r, echo=TRUE}
stem(faithful$eruptions)
```

<a href="#top">Go to top</a>

